Cross-validation of regression models: mixed model examples

### 2023-10-09 scratch/gm_addition_to_cv.Rmd
edits to gm_addition_to_cv.Rmd

Adjusting the parameters generating this data can create an example 
that illustrates problems of non-convergence with hierarchical models.
Fitted models with relatively small between-cluster variance, even simple
random intercept models, are prone to non-convergence due to the 
nature of optimizing algorithms used in `lme` and `lmer`.

In some contexts, non-convergence,
can be viewed as a diagnostic of a poor model. However, with a simple 
random intercept model where it may indicate a small between-cluster
variance, non-convergence is an annoyance since it would be statistically correct to
report that the data are consistent with a zero between-cluster population variance,
i.e. that observed between-cluster variance in the data can be explained by 
the between-individual variability within clusters.  This leads to perfectly
valid statistical inferences. Although the calculation of accurate p-values for hypotheses concerning
random effects may be complicated technically
by the fact that the estimated variance is at the boundary of the parameter space,
the estimated variance-covariance matrix of fixed-effect parameters will generally
be non-singular and may be used for valid Wald tests of linear hypotheses 
of fixed-effect parameters.

Using cross-validation based on fitting random subsets compounds the problem
since subsets on which there is non-convergence cause the cross-validation to
fail unless non-converging subsets are omitted, which may bias results since non-convergence
is likely to be related to the estimated between-cluster variances.

By varying the parameter for between-cluster variance, we can use the approach
below to generate data sets with varying propensities to produce
non-convergence in order to explore the consequences of non-convergence and to 
compare the performance of different optimizers. **to do**

@@

Note that since the independence of clusters is a reasonable assumption
while the independence of individual observations is not, 
cluster-based cross-validation has a stronger theoretical justification.

@@

  set.seed(123)
  Nb <- 10       # number of groups
  Nw <- 5        # number of individuals within groups
  Bb <- 0        # between group regression coefficient on group mean
  SDre <- 2.0    # between group SD of random level relative to group mean of X

@@

  dd <- data.frame(
    group = factor(rep(1:Nb, each = Nw)),
    x = 20 + seq(1, Nb*Nw)
  ) %>% 
    within(
      {
        xm  <- ave(x, group, FUN = mean) # withing group mean
        y <- 10 +
          Bb * xm +                  # contextual effect
          Bw * (x - xm) +            # within-group effect
          SDre * rnorm(Nb)[group] +  # random level by group
          SDwithin * rnorm(Nb*Nw)    # random error within groups
      }
    )
}
```

@@

We next obtain six fits with three models and two fitting functions.
```{r}
# random intercept models

@@

modlist <- list(fit0, fit0l, fit1, fit1l, fit2, fit2l)
names(modlist) <- sapply(modlist, formula)
```
We can compare estimated fixed-effect coefficients.

```{r, message=FALSE,warning=FALSE}
tbl <- do.call(compareCoefs, modlist) 
tbl[] <- round(tbl,4)
tbl[is.na(tbl)] <- ''
kbl(tbl, align = 'r', caption = "Comparison of estimated fixed effects")
```

@@

```{r cross-validation}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Cross-validation using lmer with random intercept model and x and the group mean of x as predictors"

cvs_cases <- lapply(modlist, cv) %>% lapply(rbind) %>% do.call(rbind,.)
cvs_clusters <- lapply(modlist, cv, cluster = 'group') %>% lapply(rbind) %>% do.call(rbind,.)

dout <- as.data.frame( rbind( cvs_cases, cvs_clusters))
# nrow(dout)
dout$cv_method <- rep(c('cases','clusters'), each = nrow(dout)/2)
dout$model <- rep(names(modlist), 2)
# dout$model <- factor(rep(c('~ 1','~ x','~ x + xm'), nrow(dout)/3))
dout$method <- rep(c('lmer','lme'), nrow(dout)/2)
dout$CV <- as.numeric(as.character(dout$`CV crit`))
dout$model <- factor(sub(' \\+ .1 .*$','', dout$model))

dout <- dout[order(dout$model),]
head(dout)
xyplot(CV ~ model , subset(dout, method == 'lmer'), groups = cv_method,
       par.settings = list(superpose.line = list(lty=c(2,3),lwd = 3)),
       type = 'l',
       auto.key = T,
       scales = list(x = list(alternating = F), y = list(log=T)))

@@

>>>>

Adjusting the parameters generating this data can be used to create examples
that illustrates problems of non-convergence with hierarchical models.
Fitted models with relatively small between-cluster variance, even those using simple
random intercept models, are prone to non-convergence due to the
nature of optimizing algorithms used in `lme` and `lmer`.

In some contexts, non-convergence,
can be viewed as a diagnostic of a poor model. However, with a simple
random intercept model where it may indicate a small between-cluster
variance, non-convergence is an annoyance because it would be statistically correct to
report that the data are consistent with a zero between-cluster population variance,
i.e. that observed between-cluster variance in the data can be explained by
the between-individual variability within clusters.  This leads to
valid statistical inferences. Although the calculation of accurate p-values for hypotheses concerning
random effects may be complicated
by the fact that the estimated variance is at the boundary of the parameter space,
the estimated variance-covariance matrix of fixed-effect parameters will generally
be non-singular and may be used for valid Wald tests of linear hypotheses
of fixed-effect parameters.

Using cross-validation based on fitting random subsets compounds the problem
since subsets for which there is non-convergence cause the cross-validation to
fail unless non-converging subsets are omitted, which may bias results since non-convergence
is likely to be related to the estimated between-cluster variances.

Varying the parameter for between-cluster variance (`SDre` in the code below),
will generate data sets with varying propensities to produce
non-convergence. These data sets can be used to explore the consequences of non-convergence and to
compare the performance of different optimizers.

@@

Since the independence of clusters may be a reasonable assumption, as is the
the independence of individual observations from the same cluster *conditional*
on clusters, each form of resampling addresses the
suitability of models
to answer different questions.
Results based on individual resampling generalize to new samples from the same
clusters. Resampling clusters generalizes to new samples of clusters and
cases.

@@

  set.seed(123)
  Nb <- 10       # number of groups
  Nw <- 5        # number of individuals within groups
  Bb <- 0        # between-group regression coefficient on group mean
  SDre <- 2.0    # between-group SD of random level relative to group mean of X

@@

  dd <- data.frame(
    group = factor(rep(1:Nb, each = Nw)),
    x = 20 + seq(1, Nb*Nw)
  ) %>%
    within(
      {
        xm  <- ave(x, group, FUN = mean) # within-group mean
        y <- 10 +
          Bb * xm +                  # contextual effect
          Bw * (x - xm) +            # within-group effect
          SDre * rnorm(Nb)[group] +  # random level by group
          SDwithin * rnorm(Nb*Nw)    # random error within groups
      }
    )
}
```

@@

We next obtain six fits with three models and two fitting functions.
```{r}
# random intercept only models

@@

modlist <- list(fit0, fit0l, fit1, fit1l, fit2, fit2l)
names(modlist) <- c('lmer:~1','lme:~1','lmer:~x','lme:~x','lmer:~x+xm','lme:~x+xm')
modlist <- do.call(models, modlist)
```
We can compare estimated fixed-effect coefficients.

```{r, message=FALSE,warning=FALSE}
tbl <- do.call(compareCoefs, modlist)
colnames(tbl) <- names(modlist)
tbl[] <- round(tbl,4)
tbl[is.na(tbl)] <- ''
kbl(tbl, align = 'r', caption = "Comparison of estimated fixed effects")
```

@@

We carry out cross-validation resampling individuals within clusters, based on
fixed and predicted random effects, and resampling clusters, based on
fixed effects only.
In order to reduce between-model variability in comparisons of models,
the `cv` method applied to the list of models created by the `models` function,
performs cross-validation with the same training and validation sets
for each model in order to reduce the between-model variability in the cross-validation
criterion that stems from the random variation in training set selection.

```{r cross-validation}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Cross-validation using lmer and lme comparing random intercept models with varying fixed effects and with case- and cluster-based resampling."

cvs_cases <- cv(modlist, data = dd)
cvs_clusters <- cv(modlist, data = dd, cluster = 'group')

cvs_cases <- cvs_cases %>% lapply(rbind) %>% do.call(rbind,.) %>% as.data.frame
cvs_clusters <- cvs_clusters %>% lapply(rbind) %>% do.call(rbind,.) %>% as.data.frame

cvs_cases$modlist <- names(modlist)
cvs_clusters$modlist <- names(modlist)
cvs_cases$cv_method <- 'cases'
cvs_clusters$cv_method <- 'clusters'

dout <-
  within(
    as.data.frame( rbind( cvs_cases, cvs_clusters)),
    {
      model <- factor(sub('lme.*:','',modlist))
      method <- factor(sub(':.*$','',modlist))
      CV <- as.numeric(as.character(`CV crit`))
      cv_method <- factor(cv_method)
    }
)

dout <- dout[order(dout$model),]

xyplot(CV ~ model , dout, groups = cv_method:method,
       par.settings =
         list(superpose.line = list(lty=c(2,3),lwd = 3),superpose.symbol=list(pch=21:24, fill='transparent', cex =1)),
       type = 'b',
       auto.key = T,
       scales = list(y = list(log=T)))

@@

With this example, `lme` and `lmer` give similar results, but resampling cases versus resampling clusters provides
very different comparisons of models.  The model with X alone, without the contextual mean of X, is assessed as
fitting very poorly sampling clusters and, relatively, much better resampling individual cases.  This illustrates the
need to perform cross-validation at different levels.

B
B
B
B
B
B
B
B
B
B
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
A
